---
title: |
      | CS&SS/STAT 566 - Causal Modelling
      | \Large Quiz Section 6
      | PC Algorithm
output: pdf_document
---
# Preamble: Package Installation
To begin, we first need to install and load the requisite packages.
```{r,include=FALSE}
rm(list=ls())
### Load required packages
##  Separate commands because some are large and the
##  download command may be slow or fail.
##  (However, if you re-run the commands it will not reload
##   packages that are already there.)

if (!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")}
    
if (!requireNamespace("Rgraphviz", quietly = TRUE))
    BiocManager::install("Rgraphviz",update=FALSE)
library("Rgraphviz")

if (!requireNamespace("RBGL", quietly = TRUE)){
    BiocManager::install("RBGL",update=FALSE)}
library("RBGL")

if (!requireNamespace("abind", quietly = TRUE)){
    install.packages("abind",update=FALSE)}
library("abind")

if (!requireNamespace("corpcor", quietly = TRUE)){
    install.packages("corpcor",update=FALSE)}
library("corpcor")

if (!requireNamespace("sfsmisc", quietly = TRUE)){
    install.packages("sfsmisc",update=FALSE)}
library("sfsmisc")

if (!requireNamespace("robustbase", quietly = TRUE)){
    install.packages("robustbase",update=FALSE)}    
library("robustbase")

if (!requireNamespace("pcalg", quietly = TRUE)){
    install.packages("pcalg",update=FALSE)}
library("pcalg")

if (!requireNamespace("graph", quietly = TRUE)){
    install.packages("graph",update=FALSE)}
library("graph")

```

```{r}
## Boolean checking whether Rgraphviz loaded
plotcpdag <- "Rgraphviz" %in% print(.packages(lib.loc = .libPaths()[1])) 
```
# Dataset 1
Let's look at this dataset and try to get the CPDAG, first manually to demonstrate how the PC algorithm works, then at once to demonstrate the \texttt{pcalg} package.
```{r}
# Load data and some utility functions (showAmat; showEdgeList)
# (Thanks to Markus Kalisch!)
source("https://www.stat.washington.edu/tsr/s566/labs/pc-output-utilities.r")
data <- read.table("https://www.stat.washington.edu/tsr/s566/labs/lab4DAG1.dat")
attr(data,"names")
x <- data$x
y <- data$y
z <- data$z
w <- data$w

names <- attr(data,"names")
```
We can run some correlation tests
```{r}
# Correlation matrix
cor(data)
# Graphical display
pairs(data, lower.panel = NULL)
```
Another way is to look at regressions. We'd have to look at a lot of regressions though, but let's power through for this first DAG
```{r}
# Check marginal independences








# Check conditional independences, S = 1








# Check conditional independences, S = 2








```
What do you see? Tedious, isn't it?

So instead of doing that, let's run it all at once!
```{r}
##### Using the PC Algorithm to do the job for us:
data <- read.csv("C:/Users/26862/Desktop/data_RS.csv",header = TRUE)
n <- nrow(data)
p <- ncol(data)
indepTest <- gaussCItest
suffStat <- list(C=cor(data), n = n)

## estimate CPDAG
alpha <- 0.05
pc.fit <- pc(suffStat, indepTest, p = p, alpha = alpha, verbose = TRUE)
showAmat(pc.fit)

if (plotcpdag) {
  plot(pc.fit, main = "Estimated CPDAG",labels=c("RS","age","knowledge","edu","income"))
  ## Note undirected edges are represented here as  <->
}
```

Read through the output - can you see what it has done?

# Dataset 2
```{r}
data <- read.table("https://www.stat.washington.edu/tsr/s566/labs/lab4DAG2.dat")
attr(data,"names")

x <- data$x
y <- data$y
z <- data$z
w <- data$w
```

Let's run some correlation tests
```{r}
# Correlation matrix
cor(data)
# Graphical display
pairs(data, lower.panel = NULL)
```

Now check for independence!
```{r}






```
... or if you believe the PC Algorithm at this point, get it to do the job for you
```{r}
##### Using the PC Algorithm
n <- nrow(data)
p <- ncol(data)
indepTest <- gaussCItest
suffStat <- list(C=cor(data), n = n)
## estimate CPDAG
alpha <- 0.05
pc.fit <- pc(suffStat, indepTest, p = p, alpha = alpha, verbose = TRUE)
showAmat(pc.fit)
showEdgeList(pc.fit,names)
if (plotcpdag) {
  plot(pc.fit, main = "Estimated CPDAG",labels=c("x","y","z","w"))
  ## Note undirected edges are represented here as  <->
}
```

# Dataset 3
Ok, last one:
```{r}
data <- read.table("https://www.stat.washington.edu/tsr/s566/labs/lab4DAG3.dat")
attr(data,"names")

x <- data$x
y <- data$y
z <- data$z
w <- data$w

names <- attr(data,"names")
```

Maybe look at some independences
```{r}
# Some linear regressions again
summary(lm(z~w, data))$coef

summary(lm(z~w + x, data))$coef
summary(lm(z~w + y, data))$coef
```

Good enough, now for the PC Algorithm
```{r}
##### Using the PC Algorithm
n <- nrow(data)
p <- ncol(data)
indepTest <- gaussCItest
suffStat <- list(C=cor(data), n = n)

## estimate CPDAG
alpha <- 0.05
pc.fit <- pc(suffStat, indepTest, p = p, alpha = alpha, verbose = TRUE)
showAmat(pc.fit)
showEdgeList(pc.fit,names)
if (plotcpdag) {
  plot(pc.fit, main = "Estimated CPDAG",labels=c("x","y","z","w"))
}
```


```{r}
data2 <- read.csv("C:/Users/26862/Desktop/data_RB.csv",header = TRUE)
## define sufficient statistics
suffStat <- list(dm = data2, nlev = c(2,2,2,3), adaptDF = FALSE)
V <- colnames(data2)
## estimate CPDAG
pc.D <- pc(suffStat,
           ## independence test: G^2 statistic
           indepTest = disCItest, alpha = 0.01, labels = V, verbose = TRUE)
if (require(Rgraphviz)) {
  ## show estimated CPDAG
  plot(pc.D, main = "Estimated CPDAG",labels=c("RB","risk","household","city"))
}
```




















